Values Encoded in Machine Learning Research

### 2025-02-14 latex/paper.tex
restructuring of section analysis

\subsection{Keyword based analysis}
To analyze the values encoded in these papers, we adopted the 74 keywords used by Birhane et al. (2023) in their study of influential ML papers. In their paper
they write that they use 67 values, but in their annotation template for manual annotations (annotations.tsv) they include 74 values. In their results folder they
just mention 62 values. In order to search for as many values as possible we decided to proceed with the 74 values from the annotation file in their data folder.  
These keywords include terms related to performance, generalization, novelty, societal impact, and ethical considerations. 
Unlike Birhane et al., who relied on manual annotation, we developed and deployed an automated keyword-scanning algorithm to systematically identify the presence of 
these keywords across the abstracts, introductions, and conclusions of the selected papers.
The Table\ref{tab:keyword_table} in the Appendix provides a comprehensive list of the values and their associated keywords used in our analysis.
For example we did not only query for the additonal keyword ``Novelty'' but also for ``novel'' as these two
words express the same value. 
For each keyword occurrence, contextual information (the sentences with the keyword occurence) were extracted to ensure accurate interpretation. 
This semi-automated process balanced the scale of automated analysis with the depth of manual review. One major limitation to this keyword based approach is 
that for some keywords this works better than for others and these keywords might then be under- or overrepresented in the analysis. For example for the keyword ``Fairness'' is really likely that the 
keyword search will find the occurences if the paper adresses this topic because they will most liekly use the word in the context. 
As a counter example the keyword `Easy to implement'' might not be found as easily because the word ``easy'' is used in many other contexts as well and therefore we decided not to query for it in order to avoid that sentences are flagged with 
the value that do not contain it.
There are many different words in which the researchers could express that their model is easy to implement and they are hard to capture all by a simple keyword search.
Even tough the keyword search is not perfect, it is a good way to get a first overview of the values encoded in the papers and to get a first impression of the values. To stay in the scope of this seminar paper 
we therefore decided to use this approach but we want to explicitly mention this limitation.

\section{Results}
\subsection{Quantitative Summary}
We measure the prevelence (keyword present in a paper at least once) of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference. First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
and lastly we compare the results from the 2023 papers with the results from the 2008/09 and 2018/19 papers from Birhane et al. \cite{valuesInML2021}. 
In line with the Birhane et al. paper we highlight the values that fall under `User rights'. These are `Interpretable (To Users)', `Deferral To Humans', `Privacy', `User influence', `Not Socially Biased' and `Fairness'. 
These values are highlighted in all plots in orange. The second group of values highlighted are values belonging to `Ethical Principles', These keywords are `Beneficence', `Non-Maleficence', `Respect for Law and Public Interest', 
`Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.
\subsubsection{ICML and Neurips 2023}
In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (99\%), accuracy (85\%), 
Understanding (for researchers) (82.5\%), Efficiency (81\%) and Fast (80\%). Among the values related to user rights and stated in 
ethical principles the most common ones are `Explicability' (22.5\%) and Privacy (21.5\%).

@@

By analysing the 100 most influencial paper from both conferences seperately we can see in Figure \ref{fig:percentage_comparison_conferences}
that the distribution of values are quite similar in both conferences. The values performance and accuracy are among the top 5 values for both conferences. 
One can consider that the maximum length for a paper is nine pages for Neurips\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
This different limitation in the maximum length of the paper does not seem to have a big impact on the distribution of values in the paper.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{../plots/percentage_comparison_conferences.png}
    \caption{Comparison of prevalence of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023. Values related to user rights are highlighted in orange, and values belonging to ethical principles are highlighted in blue.}
    \label{fig:percentage_comparison_conferences}
\end{figure}

@@

\subsubsection{Comparison 2008/09 and 2018/19 to 2023}
In Figure \ref{fig:percentage_comparison_years} we compare the distribution of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023 with the results from Birhane et al.
There are a few values that have a high percentage in both the 2023 papers and the papers from 2008/09 and 2018/19. 
These are performance (2023: 99\% and Brihane: 96\%), generalisation(77\% and 89\%) and efficiency (81\% and 84\%). 
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{../plots/percentage_comparison_brihane.png}
    \caption{Comparison of prevelance of values from 2023 to 08/09 and 18/19 from Brihane et al.}
    \label{fig:percentage_comparison_years}
\end{figure}
Strong differences as in values such as `Automatic' and `Progress' can potentially be explaind due to the fact that these 
words occur often also without specific meaning towards the values.

Zooming onto the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 except for `Deferral to humans' and `User influence'. For the latter the 
decrease is so small that it could also be due to the keyword search. Especially in the values `Privacy' and `Explicability' we see a strong increase. 
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth,height=0.32\textheight]{../plots/subset_percentage_comparison.png}
    \caption{Comparison of values related to user rights and to ethical principles.}
    \label{fig:subset_comparison_years}
\end{figure}

\subsection{Qualitative Summary}
In order to investigate if the keyword search was able to capture values correctly, we manually review 
the sentences flagged for specific values. This qualitative analysis provides a nuanced understanding of how values are encoded in the papers. 
Table \ref{tab:qualitative_examples_user_influence} provides examples of sentences from the papers that reflect the value `User influence'. These
examples show that by the automated keywords search in this case sentences get flagged that actually stand in context with the value. 
\begin{table}[H]
    \centering
    \begin{tabular}{p{12cm}}
        \toprule
        ``However, in contrast to existing works that target a specific application, without a well defined objective, we propose a more general approach that allows us to unify different user control inputs in a more principled manner.'' \\
        \midrule
        ``Controllable Diffusion Models are designed to enable T2I diffusion models to accept more user controls for guiding the generation results.'' \\
        \midrule
        ``Recently, a surge of methods have been proposed to gain wider and better user controllability.'' \\
        \bottomrule
    \end{tabular}
    \caption{Random examples of the value \textit{User influence} in the context.}
    \label{tab:qualitative_examples_user_influence}
\end{table}

For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly in the first two example sentences.
In the last case the automated search also flaged a reference cited in the paper that was published at the ACM Conference on Fairness, Accountability, and Transparency. 
In this case the value `Fairness' was not part of the paper but it got flagged. This shows a limitation of the automated approach. One could argue that citing a paper that deals with fairness
also coud be counted as a value in the paper.

\begin{table}[H]
    \centering
    \begin{tabular}{p{12cm}}
        \toprule
        ``In scenario ( 1), Table 4 shows the fairness issues of GPT-3.5 and GPT-4.'' \\
        \midrule
        ``It is unclear whether other properties that have been extensively studied in the literature for the existing public BERT checkpoint, such as robustness, out-of-distribution generalization, fairness or inherent biases, carry over to the crammed model.'' \\
        \midrule
        ``In 2022 ACM Confer- ence on Fairness, Accountability, and Transparency , pp.'' \\
        \bottomrule
    \end{tabular}
    \caption{Random examples of the value \textit{Fairness} in the context.}
    \label{tab:qualitative_examples_fairness}
\end{table}

@@

\section{Discussion}
\label{sec:discussion}
Our study extends the analysis of values encoded in machine learning (ML) research by examining the most influential papers from NeurIPS 2023 and ICML 2023. In alignment with prior work by Birhane et al. (2023), 
we find that performance, accuracy, generalization, and efficiency remain dominant values in contemporary ML research. However, our findings also suggest an increasing emphasis on ethical considerations, particularly 
explicability and privacy, indicating a growing awareness of the societal implications of ML advancements.
A key observation in our study is that the priorities of ML research largely remain entrenched in technical performance metrics. The emphasis on state-of-the-art (SOTA) performance and efficiency continues to drive the field, 
yet the explicit inclusion of values related to fairness, transparency, and social responsibility remains relatively limited. While there is a slight increase in the prevalence of values associated with user rights, such as 
privacy and explicability, these considerations are still secondary to performance-driven objectives. This reinforces concerns raised by Birhane et al. (2023) that ML research prioritizes technical achievements over broader 
social impact.
The keyword-based methodology we employed, while effective for large-scale analysis, introduces limitations regarding contextual accuracy. Automated searches for ethical principles may fail to capture nuanced discussions, 
as seen in cases where references to fairness or accountability appeared only in citations rather than substantive discourse. This underscores the need for complementary qualitative analysis to provide deeper insights into 
the way these values are operationalized in ML research.

@@

\section{Conclusion}
Our analysis reveals that while ML research is increasingly acknowledging ethical concerns such as privacy and explicability, technical performance metrics continue to dominate. 
The persistence of values such as performance, efficiency, and generalization, while important for technological progress, indicates that broader social considerations remain underprioritized. 
Despite the growing discourse around fairness, transparency, and ethical AI, these values are not yet as deeply embedded in ML research as one might hope.

This study underscores the need for a more deliberate integration of ethical principles into ML research. Future work should explore alternative evaluation metrics that balance technical performance 
with social responsibility. Moreover, institutions and funding bodies should incentivize research that explicitly addresses societal needs, ensuring that ML advancements contribute to equitable and 
inclusive technological progress.

@@

    Promising & \\ 

@@

    Performance & \\ 

@@

    Effectiveness & \\ 
    Successful & \\ 

@@

    Improvement & \\ 
    Progress & \\ 

@@

    Useful & \\ 

@@

    Beneficence & beneficable, welfare, positive impact, well-being, prosocial \\ 

>>>>

\subsection{Keyword based analysis}
To analyze the values encoded in these papers, we adopted the 74 keywords used by Birhane et al. (2023) in their study of influential ML papers. In their paper
they write that they use 67 values, but in their annotation template for manual annotations (annotations.tsv) they include 74 values. In their results folder they
just mention 62 values. In order to search for as many values as possible we decided to proceed with the 74 values from the annotation file in their data folder.  
These keywords include terms related to performance, generalization, novelty, societal impact, and ethical considerations. 
Unlike Birhane et al., who relied on manual annotation, we developed and deployed an automated keyword-scanning algorithm to systematically identify the presence of 
these keywords. Brihane et al. did only analyse the sections abstract, introduction, discussion and conclusion of the selected papers. We assume that they focused on these parts of the papers because they are most likely to contain the values and to minimize the workload for the manual annotations. Because we used an automated approach we decided to scan the whole paper (excluding the References section). We exclude this section because otherwise here references get flagged with the value `Fairness' for example with they were pubslished at the ACM Conference on Fairness, Accountability, and Transparency. This would result in a false positive.

The Table\ref{tab:keyword_table} in the Appendix provides a comprehensive list of the values and their associated keywords used in our analysis.
For example we did not only query for the additonal keyword ``Novelty'' but also for ``novel'' as these two
words express the same value. 
For each keyword occurrence, contextual information (the sentences with the keyword occurence) were extracted to ensure accurate interpretation. 
This semi-automated process balanced the scale of automated analysis with the depth of manual review. One major limitation to this keyword based approach is 
that for some keywords this works better than for others and these keywords might then be under- or overrepresented in the analysis. For example for the keyword ``Fairness'' is really likely that the 
keyword search will find the occurences if the paper adresses this topic because they will most liekly use the word in the context. 
As a counter example the keyword `Easy to implement'' might not be found as easily because the word ``easy'' is used in many other contexts as well and therefore we decided not to query for it in order to avoid that sentences are flagged with 
the value that do not contain it.
There are many different words in which the researchers could express that their model is easy to implement and they are hard to capture all by a simple keyword search.
Even tough the keyword search is not perfect, it is a good way to get a first overview of the values encoded in the papers and to get a first impression of the values. To stay in the scope of this seminar paper 
we therefore decided to use this approach but we want to explicitly mention this limitation.

\section{Results}
\subsection{Quantitative Summary}
We measure the prevelence (keyword present in a paper at least once) of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference. First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
and lastly we compare the results from the 2023 papers with the results from the 2008/09 and 2018/19 papers from Birhane et al. \cite{valuesInML2021}. 
In line with the Birhane et al. paper we highlight the values that fall under `User rights'. These are `Interpretable (To Users)', `Deferral To Humans', `Privacy', `User influence', `Not Socially Biased' and `Fairness'. 
These values are highlighted in all plots in orange. The second group of values highlighted are values belonging to `Ethical Principles', These keywords are `Beneficence', `Non-Maleficence', `Respect for Law and Public Interest', 
`Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.

\subsubsection{ICML and Neurips 2023}
In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (97.5\%), accuracy (75\%), 
State-of-the-art (71\%), Efficiency (70\%) and Improvement (67.5\%). Among the values related to user rights and stated in 
ethical principles the most common ones are Privacy (16.5\%) and `Explicability' (15.5\%).

@@

By analysing the 100 most influencial paper from both conferences seperately we can see in Figure \ref{fig:percentage_comparison_conferences}
that the distribution of values are quite similar in both conferences. The values `Performance', `Accuracy' and `State-of-the-art' are among the top 5 values for both conferences. 
One can consider that the maximum length for a paper is nine pages for Neurips\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
This different limitation in the maximum length of the paper does not seem to have a big impact on the distribution of values in the paper.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{../plots/percentage_comparison_conferences.png}
    \caption{Comparison of prevalence of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023. Values related to user rights are highlighted in orange, and values belonging to ethical principles are highlighted in blue.}
    \label{fig:percentage_comparison_conferences}
\end{figure}

@@

\subsubsection{Comparison 2008/09 and 2018/19 to 2023}
In Figure \ref{fig:percentage_comparison_years} we compare the distribution of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023 with the results from Birhane et al.
There are a few values that have a high percentage in both the 2023 papers and the papers from 2008/09 and 2018/19. 
These are `Performance' (2023: 97.5\% and Brihane: 96\%), `Generalisation' (65.5\% and 89\%) and `Efficiency' (70\% and 84\%). 
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{../plots/percentage_comparison_brihane.png}
    \caption{Comparison of prevelance of values from 2023 to 08/09 and 18/19 from Brihane et al.}
    \label{fig:percentage_comparison_years}
\end{figure}
Strong differences as in values such as `Automatic' and `Progress' can potentially be explaind due to the fact that these 
words occur often also without specific meaning towards the values.

Zooming onto the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 except for `Interpretable (to users)' (minor decrease), `Beneficence' (strong decrease), `Deferral to humans' (low decrease) and `User influence' (minor decrease). For the values expect `Beneficence' the decrease is so small that it could be due to the keyword search. The decrease for `Beneficence' is really strong from 13\% to only 1.5\%. One possible explanation is that it is hard to query by a keyword search for this value. Especially in the values `Privacy' (from 3\% to 16.5\%), `Explicability' (from 2\% to 15.5\%) and `Fairness' (from 1\% to 10\%) we see a strong increase. 
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth,height=0.32\textheight]{../plots/subset_percentage_comparison.png}
    \caption{Comparison of values related to user rights and to ethical principles.}
    \label{fig:subset_comparison_years}
\end{figure}

\subsection{Qualitative Summary}
In order to investigate if the keyword search was able to capture values correctly, we manually review 
the sentences flagged for specific values. This qualitative analysis provides a nuanced understanding of how values are encoded in the papers. 
Table \ref{tab:qualitative_examples_user_influence} provides examples of sentences from the papers that reflect the value `User influence'. These
examples show that by the automated keywords search in this case sentences get flagged that actually stand in context with the value. 
\begin{table}[H]
    \centering
    \begin{tabular}{p{12cm}}
        \toprule
        ``However, in contrast to existing works that target a specific application, without a well defined objective, we propose a more general approach that allows us to unify different user control inputs in a more principled manner.'' \\
        \midrule
        ``However, user controllability of the generated image, and fast adaptation to new tasks still remains an open challenge, currently mostly addressed by costly and long retraining and fine-tuning or ad-hoc adaptations to specific image generation tasks.'' \\
        \midrule
        ``Recently, a surge of methods have been proposed to gain wider and better user controllability.'' \\
        \bottomrule
    \end{tabular}
    \caption{Random examples of the value \textit{User influence} in the context.}
    \label{tab:qualitative_examples_user_influence}
\end{table}

For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly 
in the first two example sentences.In the last case the automated search also flaged a reference cited in the paper that was published at the 
ACM Conference on Fairness, Accountability, and Transparency. Here the References section was wrongly inclded in the search for this paper, because
the automatic exclusion of this section does not work perfectly. In this case the value `Fairness' was not part of the paper but it got flagged. 
This shows a limitation of the automated approach. One could argue that citing a paper that deals with fairnessalso coud be counted as a value in the paper.

\begin{table}[H]
    \centering
    \begin{tabular}{p{12cm}}
        \toprule
        ``In scenario ( 1), Table 4 shows the fairness issues of GPT-3.5 and GPT-4.'' \\
        \midrule
        ``It is unclear whether other properties that have been extensively studied in the literature for the existing public BERT checkpoint, such as robustness, out-of-distribution generalization, fairness or inherent biases, carry over to the crammed model.'' \\
        \midrule
        ``In 2022 ACM Conference on Fairness, Accountability, and Transparency , pp.'' \\
        \bottomrule
    \end{tabular}
    \caption{Random examples of the value \textit{Fairness} in the context.}
    \label{tab:qualitative_examples_fairness}
\end{table}

@@

\section{Discussion}
\label{sec:discussion}
Our study extends the analysis of values encoded in machine learning (ML) research by examining the most influential papers from NeurIPS 2023 and ICML 2023. In alignment with prior work by Birhane et al. (2023), 
we find that performance, accuracy, state-of-the-art, and efficiency remain dominant values in contemporary ML research. However, our findings also suggest an increasing emphasis on ethical considerations, particularly 
explicability, privacy, and fairness indicating a growing awareness of the societal implications of ML advancements.
A key observation in our study is that the priorities of ML research largely remain entrenched in technical performance metrics. The emphasis on state-of-the-art (SOTA) performance and efficiency continues to drive the field, 
yet the explicit inclusion of values related to fairness, transparency, and social responsibility remains relatively limited. While there is a slight increase in the prevalence of values associated with user rights, such as 
privacy and explicability, these considerations are still secondary to performance-driven objectives. This reinforces concerns raised by Birhane et al. (2023) that ML research prioritizes technical achievements over broader 
social impact.
The keyword-based methodology we employed, while effective for large-scale analysis, introduces limitations regarding contextual accuracy. Automated searches for ethical principles may fail to capture nuanced discussions, 
as seen in cases where references to fairness or accountability appeared only in citations rather than substantive discourse. This underscores the need for complementary qualitative analysis to provide deeper insights into 
the way these values are operationalized in ML research.

@@

\section{Conclusion}
Our analysis reveals that while ML research is increasingly acknowledging ethical concerns such as privacy and explicability, technical performance metrics continue to dominate. 
The persistence of values such as performance, efficiency, and generalization, while important for technological progress, indicates that broader social considerations remain underprioritized. 
Despite the growing discourse around fairness, transparency, and ethical AI, these values are not yet as deeply embedded in ML research as one might hope. 
This study underscores the need for a more deliberate integration of ethical principles into ML research.

This paper analyses papers from the two big ML conferences ICML and Neurips. There are specialized ML conferences dedicated to the study of fairness, accountability and transparency in AI. It can be debated whether the fact that these specialized conferences exist ensures that a focus on performance values at ICML and Neurips seems justifiable.

@@

    Promising & - \\ 

@@

    Performance & - \\ 

@@

    Effectiveness & - \\ 
    Successful & - \\ 

@@

    Improvement & - \\ 
    Progress & - \\ 

@@

    Useful & - \\ 

@@

    Beneficence & beneficable, welfare, positive impact, well-being, prosocial, altruistic, altruism, social good, ethical principle \\ 
